Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-14 latex/paper.tex
added first part of parallelization section

\section{Preliminaries}

@@

\subsubsection{Algorithm}

\begin{algorithm}
\caption{Parallel Rank-2 NMF}
\label{alg:parrank2nmf}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
		\State Initialize local $\M[\hat]{W}$ randomly
		\While{not converged}
			\State \Comment{Compute $\M{H}$}
			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
			\State $\M[\hat]{B} = \M[\hat]{W}^\Tra \M[\hat]{A}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
			\State \Comment{Compute $\M{W}$}
			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
		\EndWhile
	\EndFunction
	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Analysis}

>>>>

\section{Preliminaries}
\label{sec:prelim}

@@

In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for the approach we take.
The running time of an algorithm is data dependent because not only does each Rank-2 NMF computation require a variable number of iterations, but also the shape of the tree can vary from a balanced binary tree with $O(\log k)$ levels to a tall, flat tree with $O(k)$ levels.
For the sake of analysis, we will assume a fixed number of NMF iterations for every node of the tree and we will analyze the cost of complete levels.

The first possibility for parallelization is across the nodes of the tree, as each Rank-2 NMF split is independent.
We choose not to parallelize across nodes in the tree for two reasons.
The first reason is that while the NMF computations are independent, choosing which nodes to split may depend on global information.
In particular, when the global target is to determine $k$ leaf clusters, the nodes must be split in order of their scores, which leads to a serialization of the node splits.
This serialization might be relaxed using speculative execution, but it risks performing unnecessary computation.
If the global target is to split all nodes with sufficiently high scores, then this serialization is also avoided and node splits become truly independent.
We choose not to parallelize in this way to remain agnostic to the global stopping criterion.

The second reason is that parallelizing across nodes requires redistribution of the input data.
Given a node split by $\hat p$ processors, in order to assign disjoint sets of processors to each child node, each of the $\hat p$ processors would have to redistribute their local data, sending data for samples not in their child's set and receiving data for those in their child's set.
The communication would be data dependent, but on average, each processor would communicate half of its data in the redistribution set, which could have an all-to-all communication pattern among the $\hat p$ processors.
For a node with $\hat n$ columns, the communication cost would be at least $O(m\hat n/ \hat p)$ words, which is much larger than the communication cost per iteration of Parallel Rank-2 NMF, as we will see in \Cref{sec:analysis}.

By choosing not to parallelize across nodes in the tree, we employ all $p$ processors on each node, and split nodes in sequence.
The primary computations used to split a node are the Rank-2 NMF and the score computation, which is based on an approximation of the largest singular value.
We use an alternating-updating algorithm for Rank-2 NMF as described in \Cref{sec:prelim}, and we parallelize it following the methodology proposed in \cite{EH+19-TR} and presented in \Cref{alg:parrank2nmf}.

The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principle singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.

\subsubsection{Algorithms}

\paragraph{\emph{Parallel Rank-2 NMF}}

\begin{algorithm}
\caption{Parallel Rank-2 NMF}
\label{alg:parrank2nmf}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
		\State Initialize local $\M[\hat]{W}$ randomly
		\While{not converged}
			\State \Comment{Compute $\M{H}$}
			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
			\State \Comment{Compute $\M{W}$}
			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
		\EndWhile
	\EndFunction
	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
\end{algorithmic}
\end{algorithm}

\paragraph{\emph{Parallel Power Method}}

\begin{algorithm}
\caption{Parallel Power Method}
\label{alg:parpowmeth}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
		\State Initialize $\V{v}$ randomly and redundantly
		\While{not converged}
			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
			\State $\sigma = \|\V{v}\|$
			\State $\V{v} = \V{v}/\sigma$
		\EndWhile
	\EndFunction
	\Ensure{$\sigma \approx \sigma_1(\M{A})$ is  redundantly owned by all procs}
\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Analysis}
\label{sec:analysis}

\paragraph{\emph{Parallel Rank-2 NMF}}

\paragraph{\emph{Parallel Power Method}}

\paragraph{\emph{Hierarchical Clustering}}
